Automatically proving equivalence by type-safe reflection

### 2016-01-18 paper-JFP-DTP/content/1-introduction.tex
section 1 : updating the mismatching indices

We finally define the function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc : Binary w x $\rightarrow$ Binary w y $\rightarrow$ Bit c $\rightarrow$ Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{lstlisting}
\end{center}
\caption{Addition of two binary numbers and a carry}
\figrule
\end{figure}

Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] but we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\]
The problem is even worse for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but it also generates two proof obligations, one for each pattern, and both of them are saying that the provided term can be converted into a term which has the expected type.
For both of these two patterns, it is in fact possible to prove the equality between the expected and the provided types, precisely because this function is correct. These proofs are made of a series of rewriting, and each rewriting uses a property about the operation $+$ on natural numbers.
For example, the proof for the second pattern looks like this :
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{lstlisting}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\figrule
\end{figure}

Thus, this kind of proofs consists of a potentially long sequence of rewriting, where each rewriting step uses one of the available properties. Without some specific automation, this sequence of rewriting is done by the user of the proof assistant. This is time consuming, and just a little change in the left or the right hand side of the equality will most of the time invalidate completely the proof done by hand. This mean that just a little change in our datatypes, or in the definitions of $addBit$ and $adc$ will require us to replace this proof by another one doing the same job, which is assuring the correspondence between the expected index and the given index. The re-usability of this kind of proofs is therefore null, because they are performing rewritings for very specific terms.

If we look at the proof we've written by hand, we realize that we're only using the existence of neutral elements (this is the lemma $plusZeroRightNeutral$), the associativity and commutativity of $+$. Thus, we're rewriting a term by using the properties of a commutative monoid.

>>>>

We finally define the function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc : Binary w x $\rightarrow$ Binary w y $\rightarrow$ Bit c $\rightarrow$ Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (vCarry0 ** (vLsb ** (carry0, lsb, _))) = addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{lstlisting}
\end{center}
\caption{Addition of two binary numbers and a carry}
\figrule
\end{figure}

Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] but we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\]
The problem is even worse for the second case where the expected index is 
\[(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\] while we're trying to provide a term index over 
\[plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but it also generates two proof obligations, one for each pattern, and both of them are saying that the provided term can be converted into a term which has the expected type.
For both of these two patterns, it is in fact possible to prove the equality between the expected and the provided types, precisely because this function is correct. These proofs are made of a series of rewriting, and each rewriting uses a property about the operation $+$ on natural numbers.
For example, the proof for the second pattern looks like this :
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{lstlisting}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\figrule
\end{figure}

Thus, this kind of proofs consists of a potentially long sequence of rewriting, where each rewriting step uses one of the available properties. Without some specific automation, this sequence of rewriting is done by the user of the proof assistant. This is time consuming, and just a little change in the left or the right hand side of the equality will most of the time invalidate completely the proof done by hand. This mean that just a little change in our datatypes $Bit$ and $Binary$, or in the definitions of $addBit$ and $adc$ will require us to replace this proof by another one doing the same job, which is assuring the correspondence between the expected index and the given index. The re-usability of this kind of proofs is therefore null, because they are performing rewritings for very specific terms.

If we look at the proof we've written by hand, we realize that we're only using the existence of neutral elements (this is the lemma $plusZeroRightNeutral$), and the associativity and commutativity of $+$. Thus, we're rewriting a term by using the properties of a commutative monoid.
